Basics of cubic spline models

### 2021-11-24 text/21/cubic-splines.Rmd
More ggplot2 work with cubic-splines.Rmd

```{r}
library(ggplot2)
x <- 0:20
y <- rep(40, 21)
y[6:21] <- y[6:21]-3*(1:15)
y[11:21] <- y[11:21]+7*(1:10)
y[16:21] <- y[16:21]-(1:5)^2
y <- y+round(rnorm(10), 1)
simulated_example <- data.frame(x, y)
simulated_example
ggplot(simulated_example, aes(x, y)) +
  geom_point()
```

@@

```{r}
x <- c(0:20, setdiff(seq(0, 20, by=1/64), 0:20))
xm <- data.frame(
  c1 =rep(1, length(x)),
  c2 =x,
  c3 =x^2,
  c4 =x^3,
  c5 =(x> 5)*rep(1, length(x)),
  c6 =(x> 5)*(x- 5),
  c7 =(x> 5)*(x- 5)^2,
  c8 =(x> 5)*(x- 5)^3,
  c9 =(x>10)*rep(1, length(x)),
  c10=(x>10)*(x-10),
  c11=(x>10)*(x-10)^2,
  c12=(x>10)*(x-10)^3,
  c13 =(x>15)*rep(1, length(x)),
  c14=(x>15)*(x-15),
  c15=(x>15)*(x-15)^2,
  c16=(x>15)*(x-15)^3,
  y=NA
)
xm$y[1:21] <- y
```

```{r}
kable(xm[1:21, 1:4])
single_cubic_model <- lm(y~c2+c3+c4, data=xm)
single_cubic_model %>% 
  augment(newdata=xm) %>%
  ggplot(aes(x, y)) + 
    geom_point() +
    geom_line(aes(y=.fitted)) +
    ggtitle("Single cubic fit to the data")
```

@@

```{r}
kable(xm[1:20, ])
xm$y <- NA
xm$y[1:20] <- y
four_independent_cubics <- 
  lm(y~c2+c3+c4+c5+c6+c7+c8+c9+c10+c11+c12+c13+c14+c15+c16, data=xm)
augment(four_independent_cubics, newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y)) + 
    ggtitle("Discontinous cubic splines")
```

@@

```{r}
kable(xm[1:20, c(1:4, 6:8, 10:12, 14:16)])
four_continuous_cubics <- 
  lm(y~c2+c3+c4+c6+c7+c8+c10+c11+c12+c14+c15+c16, data=xm)
augment(four_continuous_cubics, newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y)) +
    ggtitle("Continous but not smooth cubic splines")

@@

```{r}
kable(xm[1:20, c(1:4, 7:8, 11:12, 15:16)])
four_smooth_cubics <- 
  lm(y~c2+c3+c4+c7+c8+c11+c12+c15+c16, data=xm)
augment(four_smooth_cubics, newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y)) + 
    ggtitle("Continous and Smooth (1st derivative) splines")

@@

```{r}
kable(xm[1:20, c(1:4, 8, 12, 16)])
spline_model <- 
  lm(y~c2+c3+c4+c8+c12+c16, data=xm)
augment(spline_model, newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y)) +
    ggtitle("Continous and smooth (1st and 2nd derivatives) cubic splines")
```

@@

```{r}
linear_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=1, intercept=TRUE)
kable(linear_spline[1:21, ])
plot(x[1:21], linear_spline[1:21 , 1])
plot(x[1:21], linear_spline[1:21 , 2])
plot(x[1:21], linear_spline[1:21 , 3])
plot(x[1:21], linear_spline[1:21 , 4])
plot(x[1:21], linear_spline[1:21 , 5])

lm(xm$y~linear_spline) %>%
  augment(newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y)) + 
    ggtitle("Linear splines")
```

@@

lm(xm$y~quadratic_spline) %>%
  augment(newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y)) + 
    ggtitle("quadratic splines")
```

@@

lm(xm$y~cubic_spline) %>%
  augment(newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y)) + 
    ggtitle("cubic splines")
```

>>>>

```{r}
library(ggplot2)
x <- 0:20
y <- rep(40, 21)
y[6:21] <- y[6:21]-3*(0:15)
y[11:21] <- y[11:21]+7*(0:10)
y[16:21] <- y[16:21]-(0:5)^2
y <- y+round(rnorm(21), 1)
simulated_example <- data.frame(x, y)
simulated_example
ggplot(simulated_example, aes(x, y)) +
  geom_point()
```

@@

```{r}
x <- c(0:20, setdiff(seq(0, 20, by=1/64), 0:20))
```

You can get a variety of splines by defining constant, linear, quadratic, and cubic terms and then shift those functions to the right. After shifting, fill in the hole to the left with zeros.

```{r}
xm <- data.frame(
  c1 =rep(1, length(x)),
  c2 =x,
  c3 =x^2,
  c4 =x^3,
  c5 =(x> 5)*rep(1, length(x)),
  c6 =(x> 5)*(x- 5),
  c7 =(x> 5)*(x- 5)^2,
  c8 =(x> 5)*(x- 5)^3,
  c9 =(x>10)*rep(1, length(x)),
  c10=(x>10)*(x-10),
  c11=(x>10)*(x-10)^2,
  c12=(x>10)*(x-10)^3,
  c13=(x>15)*rep(1, length(x)),
  c14=(x>15)*(x-15),
  c15=(x>15)*(x-15)^2,
  c16=(x>15)*(x-15)^3,
  y=NA
)
xm$y[1:21] <- y
```

```{r}
display_terms <- function(df, x, y, lb, z=y) {
  full_data <- data.frame(x=df[ , x], y=df[ , y], z=df[ , z])
  integer_data <- full_data[1:21, ]
  ggplot(full_data, aes(x, y)) +
      ggtitle(lb) +
      geom_line() +
      geom_point(aes(x, z), na.rm=TRUE) +
      xlab(" ") +
      ylab(" ")
}

lb1 <- rep(c("Intercept", "Linear term", "Quadratic term", "Cubic term"), 4)
lb2 <- rep(c("for the full range", "restarted at x=5", "restarted at x=10", "restarted at x=15"), each=4)
lb <- paste(lb1, lb2)
for (j in 1:16) {
  xm %>% display_terms("c2", paste0("c", j), lb[j]) -> cplot
  plot(cplot)
}
```

### Single cubic polynomial

This is the matrix of independent variables for a single cubic polynomial.

```{r}
kable(xm[1:21, 1:4])
```

The graph shown below represents the best fitting single cubic polynomial.

```{r}
lm(y~c2+c3+c4, data=xm) %>%
  augment(newdata=xm) %>%
  data.frame %>%
  display_terms("c2", ".fitted", "Single cubic fit", "y")
```

### Separate cubic polynomials

@@

```{r}
kable(xm[1:20, ])
```

```{r}
four_independent_cubics <- 
  lm(y~c2+c3+c4+c5+c6+c7+c8+c9+c10+c11+c12+c13+c14+c15+c16, data=xm)
augment(four_independent_cubics, newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    # geom_point(xm[1:21, ], aes(c2, y)) +
    ggtitle("Discontinous cubic splines")
```

@@

```{r}
kable(xm[1:20, c(1:4, 6:8, 10:12, 14:16)])
four_continuous_cubics <- 
  lm(y~c2+c3+c4+c6+c7+c8+c10+c11+c12+c14+c15+c16, data=xm)
augment(four_continuous_cubics, newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    # geom_point(xm[1:21, ], aes(c2, y)) +
    ggtitle("Continous but not smooth cubic splines")
```

@@

```{r}
kable(xm[1:20, c(1:4, 7:8, 11:12, 15:16)])
four_smooth_cubics <- 
  lm(y~c2+c3+c4+c7+c8+c11+c12+c15+c16, data=xm)
augment(four_smooth_cubics, newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y), na.rm=TRUE) +
    ggtitle("Continous and Smooth (1st derivative) splines")

@@

```{r}
kable(xm[1:20, c(1:4, 8, 12, 16)])
spline_model <- 
  lm(y~c2+c3+c4+c8+c12+c16, data=xm)
augment(spline_model, newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    # geom_point(xm[1:21, ], aes(c2, y)) +
    ggtitle("Continous and smooth (1st and 2nd derivatives) cubic splines")
```

@@

```{r}
linear_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=1, intercept=TRUE)
kable(linear_spline[1:21, ])
```

```{r}
plot(x[1:21], linear_spline[1:21 , 1])
plot(x[1:21], linear_spline[1:21 , 2])
plot(x[1:21], linear_spline[1:21 , 3])
plot(x[1:21], linear_spline[1:21 , 4])
plot(x[1:21], linear_spline[1:21 , 5])

lm(xm$y~linear_spline) %>%
  augment(newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    # geom_point(xm[1:21, ], aes(c2, y)) +
    ggtitle("Linear splines")
```

@@

lm(xm$y~quadratic_spline) %>%
  augment(newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y), na.rm=TRUE) + 
    ggtitle("quadratic splines")
```

@@

lm(xm$y~cubic_spline) %>%
  augment(newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y), na.rm=TRUE) + 
    ggtitle("cubic splines")
```
